### 2015-01-07 Paper1/paper1B.tex

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{I meant to do that: determining the intentions of action in the face of disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{1,2\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Our actions often do not match our intentions when there are external disturbances. We derive a novel modeling approach for determining motor intent from a reaching motion that is disturbed by an unexpected force. We first show how to mathematically invert both feedforward (predictive) and feedback controls to obtain an intended trajectory. We next examine the model’s sensitivity to parameter uncertainty, and test the approach using human reaching data. We found that sensitivity to parameter inaccuracies was 1.6 $\mathrm{mm}^2$, and hence lower than the inaccuracies seen in actual human motions. Further, the largest sensitivity arose mainly from uncertainty in our assumptions of joint stiffness. We also tested the hypothesis that this would predict straight-line intent in human targeted reaching movements, even when disturbed by various pulses of force. As expected, the intended trajectory showed no change from undisturbed reaching for 300 ms after the disturbance onset. Beyond this point, however, we detected a change in intent, even when the intention was near the target.  Knowing such an intent signal is broadly applicable in areas such as in enhanced human-machine interaction, the study of impaired intent in neural disorders, the real-time determination (and manipulation) of error in training, and in any complex system that embodies planning, such as brain machine interfaces, team sports, crowds, or swarms. In addition, observing intent as it changes might act as a window into the mechanisms of planning, correction, and learning.

@@

\section*{Introduction}
There are many situations, such as disturbances, distractions, and pathologies, that interfere with execution of an intended movement. One can assume that a person who is familiar with a particular task will try to do the same thing each time, provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Challenges arise in human-machine interactions because intention and changes in intention cannot be easily discerned. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as crowd control, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

It is important to distinguish intended trajectory from \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete the experiment in a timely fashion and are usually explicitly provided targets to be reached. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action (i.e, the intended trajectory), even when the actual movement is disturbed and hence no longer matches the intent.

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle's equilibrium point. Supporters of the ``$\lambda$ model''\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement even though it did not compensate for the dynamics of the arm as a feedforward controller would. Upon Gomi and Kawato's inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement. We propose a revised and more general method, \textit{intent extraction}, which calculates the intended trajectory even in the face of disturbances by relating environmental interaction and state to intent. This method affords new ways to 1) test and falsify candidate models of motor control, 2) study the timing and composition of alterations to intent during movement, and 3) understand how intent might be altered by pathology after neural insult. This method describes a class of filters that map a disturbance, process, actuator, and realized trajectory to an intended trajectory. In human reaching, these initial ingredients are available: the process is a dynamic model of the arm, the actuators (muscles) are assumed to act as viscoelastic torque generators, the human interface forces are measured by a sensor, and the realized trajectory is measured.

Here we develop and analyze the mathematics underlying intent extraction, a method to determine one's intended trajectory. We focused, for simplicity in this initial study, on planar targeted arm reaching actions. First, we produce synthetic reaching data using a forward model so that we can examine the quality and uncertainty of intent extraction. We then explore sensitivity and failure modes, demonstrating the accuracy and utility of this method. We show efficacy on a preliminary dataset of healthy reaching movements disrupted by occasional, unpredictable disturbances. Finally, we demonstrate the viability of this method for providing a better understanding of what a person means to do when they move and are disturbed, showing how one's intended trajectory may change following a disturbance.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}
The sections below describe the theoretical method and the means by which we evaluate it. First, we considered an idealized system using synthetic data that demonstrated the concept and provided an understanding of the computational process. Next, we conducted an experiment on humans to evaluate the process using human subject data.

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but its feedback component lacks the complexity and accuracy of more recent approaches. Nevertheless, later we will discuss how this approach can be used on more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=0
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impendance, Feedback Controller}}
\end{equation}
Some terms have hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) to indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$, otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ signifies the unknown desired trajectory that we seek to discover.

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This first involves algebraic manipulation to solve for $\ddot{q}$, and then integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated numerically using a differential equation solver to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

It is important to distinguish our intended trajectory, $q_d$, from another variable often considered, the equilibrium point of the muscle actuators. Importantly, this has been claimed to be the intended movement. To consider this, we also generalize our approach to any system with a moving intention that governs the control law. This introduces two additional layers of complexity by separating the plant (arm) into its process (dynamics) and actuator components (muscles) and considering the actuator's equilibrium, $q_e$ (sometimes represented as $\lambda$). The feedforward component must use and share this actuator and hence use the actuator's equilibrium. Without loss of generality, we represent the process $P$ and actuator $A$ as linear (or linearizable) operations in some space, $x$:
\begin{equation} \label{eq:generalPlant}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
The internal model of the process ($\hat{P}$ and $\hat{E}$) predicts system actions, external disturbances, and impedance responses in order to determine $x_e$ such that $x$ will track a desired path $x_d$.    
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
We solve this for $x_e$ and substituting into \eqref{eq:generalPlant} gives:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}

@@

While we treat both the process and actuator as dynamic models in our derivation, it is important to note that both the derivation and the technique are agnostic to the process model. Instead of a dynamic model, it could be a lookup table. For instance, a switch is well-modeled as a lookup table or threshold without any consideration of its underlying mechanism. In this case, our technique could be used to determine a person's intention to flip the switch and how it changes in the face of disturbance. As long as some bidirectional relationship (even statistical) exists between state and outcome, intended outcome can be determined. This allows determination of intent in many situations of interest, even where the process is irreducibly complex.

\subsection*{Dynamic simulation of arm and intended trajectories}
While the derivation above is general, testing its application to human reaching requires choosing plant and actuator models, which themselves require physical parameters. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. For the actuator, we choose the model and parameters of Burdet et al. \cite{burdet2006stability}. We use a subject-specific constant scale factor, $c$, relative to the model of Burdet et al. \cite{burdet2006stability} to account for task-dependence of the subjects' impedance\cite{gomi1998task}.
\begin{equation}
\tau_{tp}=c(K_p\Delta q+K_d \Delta \dot{q})
\end{equation}
This uses linear regression, as in Franklin et al.\cite{franklin2003adaptation}, except that the task-perpendicular data from the first 150 ms of movement in their noise-disturbed trials is used in place of a stationary task.

Desired trajectory in time was simulated as a minimum jerk, 5th order polynomial of duration 700 ms starting and ending with zero velocity and acceleration \cite{flash1985coordination}.  In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time $500 \mathrm{\mu s}$. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

\subsection*{Experimental Design}
We chose 15 centimeters long simulated reaches of the right arm, beginning 38 centimeters out from and 5.7 centimeters left of the right shoulder then first proceeding to the right.  We used the following two types of perturbing forces for each combination of distance and direction: 
\begin{enumerate}
\item Pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
\item Noise forces began once the subject moved 3 millimeters, and lasted for the duration of the motion. The forces were drawn from a white noise generator at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10 \pi$ rad/s.
\end{enumerate}
Our error metric was deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the straight-line nominal trajectory to the target. This was evaluated across each of 150 bins measuring 1 millimeter in width and spaced evenly throughout the 15 centimeter reach. Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Mean and maximum signed error relative to the direction of disturbance, also called perpendicular deviation, were used to measure reaching accuracy in order to elucidate direction of any corrections.

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Nominal parameters values were mostly taken from Dempster \cite{dempster1955space} and Burdet et al. \cite{burdet2006stability} with typical values where subject-specific parameters (height and weight) and direct measurements where necessary (force sensor noise and drift). Standard deviations reflect expected variation in repeated measurement (ie. height and weight) or measured variance (ie. trial-to-trial variation in sensor noise or uncertainty in mass ratios reported by Dempster \cite{dempster1955space}). Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

\subsection*{Statistical Analysis}
To determine when the intended trajectory's deviation exceeded the deviation that could be explained by undisturbed reaching, we compared the maximum deviation of the undisturbed reaches to the instantaneous deviation of the pulse-disturbed reaches during 100 ms intervals. The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform one-tailed T-tests with an alpha level of 0.05. This also provided a confidence interval on the difference.

@@

\subsection*{Parameter Sensitivity}
Variance in intended trajectories due to estimated uncertainty in model parameters was lower than the natural variation in undisturbed motions. Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (Figure 1) were extracted in the presence of 220,000 variations upon the parameters (Table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $2.24 \mathrm{mm}^2$ for pulse forces and $.3 \mathrm{mm}^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 \mathrm{mm}^2$. These variances describe mean unsigned error, but do not address maximum error.

@@

These results suggested a reproducible approach for inspecting how individuals might alter their intent in response to disturbances. Actual hand paths could no longer be explained ($p=0.05$) by deviation in undisturbed movement beginning within at most 135 milliseconds after the onset of disturbance. We observed no significant change of intent during force disturbance until 300 milliseconds after the onset of disturbance (Table 2). When disturbances occurred early in reaching, intent showed some signs of correction mid-reach. When disturbances occurred later, the intent continued to the target, but interestingly, it then corrected (Figure 3 lower plots). Intention showed systematic deflections that counteracted the direction of disturbance about 200 milliseconds after the onset of the force disturbance. This counteraction did not depart from the range of undisturbed movements, however (Figure 4, green region). The movements completed with the hand returning to undisturbed levels of error within 600 milliseconds, intention differed from actual hand path for as long as one second after the onset of motion.  
\section*{Discussion}
We sought to test the suitability and robustness of an algorithm that determines a person's intent during reaching motion, even if there are disturbances. Sensitivity analysis on synthetic data revealed that errors in response to parameter variations were smaller than the trial-to-trial variance commonly observed in human reaching. Mis-estimation was largest if the stiffness was inaccurately modeled. We tested this on human reaching and found that in spite of the hand being disturbed by forces, the intended movement remained on-course to the target, perhaps changing between 200 and 300 milliseconds. This represents a new and accurate method for viewing intent and how it changes when faced with a force disturbance.

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{I Meant to Do That: Determining the Intentions of Action in the Face of Disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{1,2\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Our actions often do not match our intentions when there are external disturbances such as turbulence. We derived a novel modeling approach for determining this motor intent from targeted reaching motions that are disturbed by an unexpected force. First, we demonstrated how to mathematically invert both feedforward (predictive) and feedback controls to obtain an intended trajectory. We next examined the model's sensitivity to a realistic range of parameter uncertainties, and found that the expected inaccuracy due to all possible parameter mis-estimations was less than typical movement-to-movement variations seen when humans reach to similar targets. The largest sensitivity arose mainly from uncertainty in joint stiffnesses. Humans cannot change their intent until they acquire sensory feedback, therefore we tested the hypothesis that a straight-line intent should be evident for at least the first 120 ms following the onset of a disturbance. As expected, the intended trajectory showed no change from undisturbed reaching for more than 300 ms after the disturbance onset. Beyond this point, however, we detected a change in intent, surprisingly even when the hand is already near the target. Knowing such an intent signal is broadly applicable: enhanced human-machine interaction, the study of impaired intent in neural disorders, the real-time determination (and manipulation) of error in training, and complex systems that embody planning such as brain machine interfaces, team sports, crowds, or swarms. In addition, observing intent as it changes might act as a window into the mechanisms of planning, correction, and learning.

@@

\section*{Introduction}
Disturbances, distractions, and pathologies can interfere in many situations and prevent our actions from matching our intent. A pilot, for example, may fail to complete a maneuver because of turbulence. Such challenges are broadly present in many human-machine interactions. One may speculate on how we might use these very same machines to elucidate the underlying intention. Such a possibility would be broadly useful in any area where intended actions might be thwarted by disturbances.

It is important to distinguish intended action from \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete an experiment in a timely fashion, and are usually explicitly provided targets to reach to. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of discrete actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action (i.e, the intended trajectory), even when the actual movement is disturbed and hence no longer matches the intent.

Attempts to deduce motor intent in the past have focused on the assumed spring-like properties of human muscles. Springs produce a force according to their impedance and stretch. By measuring force, impedance, and position, Gomi and Kawato\cite{gomi1997human} were able to deduce stretch and thereby infer the muscle's equilibrium point. Supporters of the ``$\lambda$ model''\cite{feldman1995origin} hypothesized that this equilibrium point represented the intent of a movement even though it did not compensate for the dynamics of the arm as a feedforward controller would. Upon Gomi and Kawato's inspection of the equilibrium point as it evolved in time, it was clear that it was highly complex and often not anatomically realizable. Therefore, it could not well-represent the intent of a simple reaching movement.

Here we propose a revised and more general method, \textit{intent extraction}, which calculates the intended trajectory even in the face of disturbances by relating environmental interaction and state to intent. The method is a class of filters that can infer the intended trajectory from the disturbance (turbulence), process (dynamic equations), actuator (muscles), and motion. The method affords new ways to study models of motor control, understand the timing and composition of intent, and how it might be altered by disturbances and injuries. Below we first present the mathematics underlying intent extraction. Then, for simplicity in this initial study, we create synthetic reaching data to examine the quality and uncertainty of this method. Finally, we demonstrate its effectiveness extracting intended action from real human reaches that have been disrupted by unpredictable disturbances. We demonstrate this effectiveness by testing the hypothesis that disturbance does not change intended action for at least 120 milliseconds. This early experiment also provides some new insights on how intended trajectories can change in reaction to disturbances.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}
The sections below describe the theoretical method used and the means by which we evaluated it. First, we present an idealized system using synthetic data that demonstrated the concept and provided an understanding of the computational process. Next, we present an experimental study on humans that evaluated success in a simple test of the approach's ability to extract straight intentions even when disturbed.

\subsection*{Intent Extraction: Deducing the Desired Trajectory}
We first demonstrate the \textit{intent extraction} approach in human motor control, but we later show that the process is applicable to any controlled process with an invertible highest order plant term. The process begins by presuming a model of the controller. Here, we choose the well-known motion control structure of Shadmehr and Mussa-Ivaldi\cite{shadmehr1994adaptive} where the feedforward aspect of the controller perfectly predicts the plant and linearizes the system through cancellation. Additionally, linear feedback rejects position and velocity error. This model was chosen to help illuminate the approach, but later we present how to generalize this approach to more complex models without loss of generality. The equation governing the passive planar dynamics (plant) of musculoskeletal structure is of the form:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=0
\end{equation}
Where $M$ is the mass matrix, $q$ is the joint angles, $\dot{q}$ is joint angular velocity, $\ddot{q}$ is joint angular acceleration, $G$ contains both Coriolis and centripetal effects, and $E$ is any externally-applied torque. The motion behavior changes with the addition of feedforward and/or feedback controllers:
\begin{equation}
\underbrace{\overbrace{M(q)\ddot{q}}^{\text{Inertia}}+\overbrace{G(q,\dot{q})}^{\text{Coriolis, Centripal}}}_{\text{Plant}}+E=\underbrace{\overbrace{\hat{M}(q_d)\ddot{q}_d}^{\text{Inertia}}+\overbrace{\hat{G}(q_d,\dot{q}_d)}^{\text{Coriolis, Centripal}}+\hat{E}}_{\text{Feedforward Controller}}+\underbrace{K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})}_{\text{Impendance, Feedback Controller}}
\end{equation}
Some terms have hats over them ($\hat{M}$, $\hat{G}$, or $\hat{E}$) to indicate that they represent the nervous system's best estimate of the forces and dynamics it will encounter, which is also known as an internal model \cite{shadmehr1994adaptive}. This portion of the system serves as an inverse-dynamics feedforward controller that cancels out the dynamics of the arm in the torque balance. If the nervous system has sufficient experience and is expecting $E$, it is included as part of the internal model, $\hat{E}$, otherwise $\hat{E}$ is set to zero. $K_p$ and $K_d$ are the lumped impedance and feedback terms that employ a moving state equilibrium to accomplish the desired trajectory, $q_d$. This $q_d$ signifies the unknown desired trajectory that we seek to discover.

For typical dynamic simulations, in order to determine the trajectory of the system (i.e., the forward dynamics problem), this second-order differential equation is solved by numerical integration to determine the solution to the initial value problem in time. This entails algebraic manipulation to solve for $\ddot{q}$, followed by integration to determine the state trajectory. Intent determination takes the novel approach of instead solving for $\ddot{q}_d$:
\begin{equation}
\ddot{q}_d=\hat{M}(q_d)^{-1}\left\{M(q)\ddot{q}+G(q,\dot{q})+E-[\hat{G}(q_d,\dot{q}_d)+\hat{E}+K_p(q_d-q)+K_d(\dot{q}_d-\dot{q})]\right\}
\end{equation}  
$\ddot{q}_d$ can then be integrated numerically using a differential equation solver to determine the intended state trajectory $q_d(t)$. This relies on many assumptions: The model of plant and controller must be accurate and precise. The initial conditions must be available and accurate. The mass matrix estimate, $\hat{M}$, must be invertible. Externally-applied force must be precisely and accurately measured. If all of these conditions are met, then the system yields an accurate estimate of the intent.

It is important to distinguish our intended trajectory, $q_d$, from another variable often considered, the equilibrium point of the muscle actuators. Importantly, this has been claimed to be the intended movement\cite{feldman1995origin}. To consider this, we also generalize our approach to any system with a moving intention that governs the control law. This introduces two additional layers of complexity by separating the plant (arm) into its process (dynamics) and actuator components (muscles) and considering the actuator's equilibrium, $q_e$ (sometimes represented as $\lambda$). The feedforward component must use and share this actuator and hence use the actuator's equilibrium. Without loss of generality, we represent the process $P$ and actuator $A$ as linear (or linearizable) operations in some action space, $x$:
\begin{equation} \label{eq:generalPlant}
\underbrace{\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+\overbrace{\sum_{m=1}^M A_m(x-x_e)^{(m)}}^\text{Actuator}}_\text{Plant}+E=0
\end{equation}
The internal model of the process ($\hat{P}$ and $\hat{E}$) predicts system actions, external disturbances, and impedance responses in order to determine $x_e$ such that $x$ will track a desired path $x_d$.    
\begin{equation}
\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m(x_d-x_e)^{(m)}}^\text{Actuator}=0
\end{equation}
We solve this for $x_e$ and substituting into \eqref{eq:generalPlant} gives:
\begin{equation}
\overbrace{\sum_{n=1}^N P_nx^{(n)}}^\text{Process}+E=\overbrace{\sum_{n=1}^N \hat{P}_n x^{(n)}_d+\hat{E}}^\text{Internal Model}+\overbrace{\sum_{m=1}^M A_m (x_d-x)^{(m)}}^\text{Feedback (Actuator)}
\end{equation}

@@

While we treat both the process and actuator as dynamic models in our derivation, it is important to note that both the derivation and the technique are agnostic to the process model. Instead of a dynamic model, it could be a lookup table. For instance, a switch is well-modeled as a lookup table or threshold without any consideration of its underlying mechanism. In this case, our technique could be used to determine a person's intention to flip the switch and how it changes in the face of disturbance. As long as some bidirectional relationship  exists between state and outcome (even if determined empirically), intended outcome can be determined. This allows determination of intent in many situations of interest, even where the process is irreducibly complex.

\subsection*{Experimental Design}
We chose 15 centimeters long simulated reaches of the right arm, beginning at 38 centimeters out from and 5.7 centimeters left of the right shoulder and ending at 38 centimeters out from and 10.7 centimeters right of the right shoulder.  We used the following two types of perturbing forces for each combination of distance and direction: 
\begin{enumerate}
\item Pulse forces applied in one of the the two directions perpendicular to the direction of movement began when the subject had moved either 10\% or 50\% of the distance to the target and lasted for 150ms.
\item Noise forces began once the subject moved 3 millimeters, and lasted for the duration of the motion. The forces were drawn from a white noise generator at 1000 Hz with flat power spectral density of 1N, and then passed through a 4th order low-pass Butterworth filter with cutoff $10 \pi$ rad/s.
\end{enumerate}
Our error metric was deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the straight-line nominal trajectory to the target. This was evaluated across each of 150 bins measuring 1 millimeter in width and spaced evenly throughout the 15 centimeter reach. Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Mean and maximum signed error relative to the direction of disturbance, also called perpendicular deviation, were used to measure reaching accuracy in order to elucidate direction of any corrections.

\subsection*{Dynamic simulation of arm and intended trajectories}
While the derivation of intent determination is general, testing its application to human reaching requires choosing plant and actuator models, which themselves require physical parameters. Appropriate plant modeling is well-understood: measured arm segment lengths and self-reported body mass are converted into the plant's inertial, centripetal, and Coriolis terms. Anatomical landmarks and values from Dempster \cite{dempster1955space} and Winter \cite{winter2009biomechanics} relate body mass to limb mass, limb length to limb center of mass, and limb mass and length to moment of inertia. For the actuator, we choose the model and parameters of Burdet et al. \cite{burdet2006stability}. We fit subject-specific constant scale factors for body mass, $m$, and impedance, $c$, relative to this model to account for atypical mass distribution and any task-dependence of the subjects' impedance\cite{gomi1998task}.
\begin{equation}
E=c(K_p\Delta q+K_d \Delta \dot{q})+m(I(q)\Delta\ddot{q})
\end{equation}
This uses linear regression, as in Franklin et al.\cite{franklin2003adaptation}, except that task-perpendicular torque and displacement from the first 150 ms of movement from noise-disturbed trials is used in place of a stationary task.

Desired trajectory in time was idealized as a typical minimum jerk, 5th order polynomial of duration 700 ms starting and ending with zero velocity and acceleration \cite{flash1985coordination}.  In this  model, musculoskeletal stiffness was linearly related to torque. Like Burdet et al. \cite{burdet2006stability}, we approximated muscle torque by looking backward in time $500 \mathrm{\mu s}$. Note that this backward look led to negligible (sub-micrometer mean unsigned error) discrepancies in subsequent intent estimates.

\subsection*{Indices of variance-based sensitivity}
To ensure a properly spaced and efficient evaluation of sensivity, we employed Sobol-distributed matrices, generated using MATLAB's \textit{sobolset()} function and converted to parameter distributions (see Table~\ref{tab:parameters}) using inverse cumulative probability density. Nominal parameter values were mostly taken from Dempster \cite{dempster1955space} and Burdet et al. \cite{burdet2006stability} with typical values for subject-specific parameters (height and weight) and direct measurements for device-specific parameters (force sensor noise and drift). Standard deviations reflect expected variation in repeated measurement (ie. height and weight) or measured variance (ie. trial-to-trial variation in sensor noise or uncertainty in mass ratios reported by Dempster \cite{dempster1955space}). Distributions were clamped to the range of $\pm3$ standard deviations in order to limit parameters to a realistic range. Combinations and calculations were made as prescribed by Saltelli et al. \cite{saltelli2010variance} to arrive at direct and total sensitivity indices.

@@

\subsection*{Statistical Analysis}
Our principal goal was to determine if and when disturbed trajectories departed from undisturbed trajectories. To detect this departure, we use the one-tailed Student's t-test $(\alpha=.05)$ at 5 millisecond intervals to determine whether or not the deviation of the disturbed trajectories had exceeded the maximum deviations of the undisturbed trajectories. The MATLAB statistics toolbox package \cite{MATLAB:2008} was used.

@@

\subsection*{Parameter Sensitivity}
Variance in intended trajectories due to estimated uncertainty in model parameters was lower than the natural variation in undisturbed motions. Simulated point-to-point reaches disturbed by either filtered white noise forces or a pulse force perpendicular to the direction of the reach (Figure 1) were extracted in the presence of 220,000 variations upon the parameters (Table 1) according to the methods of Saltelli et al. \cite{saltelli2010variance}. Expected variance due to parameter uncertainty was $2.24 \mathrm{mm}^2$ for pulse forces and $.3 \mathrm{mm}^2$ for filtered white forces. Variance in recorded undisturbed point-to-point reaching under the same time and reach distance conditions was $2.93 \mathrm{mm}^2$. These variances describe mean unsigned error, but this uncertainty is not distributed evenly in space or time.

@@

These results suggested a reproducible approach for inspecting how individuals might alter their intent in response to disturbances. Actual hand paths could no longer be explained ($p=0.05$) by deviation in undisturbed movement within 135 milliseconds after the onset of disturbance. We observed no significant change of intent during force disturbance until 300 milliseconds after the onset of disturbance (Figure 4). When disturbances occurred early in reaching, intent showed some signs of correction mid-reach. When disturbances occurred later, the intent continued to the target, but interestingly, it then corrected (Figure 3, bottom two plots). Intention showed systematic deflections that counteracted the direction of disturbance beginning about 200 milliseconds after the onset of the force disturbance; however, this counteraction did not immediately depart from the range of undisturbed movements (Figure 4, orange region). Hand movements returned to undisturbed levels of error within 600 milliseconds, but intention did not always return to undisturbed levels within one second.

\section*{Discussion}
We sought to test the suitability and robustness of an algorithm that determines a person's intent during reaching motion, even if there are disturbances. Sensitivity analysis on synthetic data revealed that errors in response to parameter variations were smaller than the trial-to-trial variance commonly observed in human reaching. Mis-estimation was largest if the stiffness was inaccurately modeled. We tested this on human reaching and found that in spite of the hand being disturbed by forces, the intended movement remained on-course to the target, perhaps changing between 200 and 300 milliseconds. This represents a new and accurate method for viewing intent and how it changes when faced with a force disturbance.
